% The Thirteenth Day of the Ninth Moon, Year 735

\section*{The Vallaki Coup}
Around three, we set out toward Vallaki. We walked through the treelines so we'd come at it from the north, the plan being to climb over the city wall behind the church and talk with Brother Petrovich about de-wolfing us.

Along the way, we came a cross a recently dug up grave that looked as if someone had burst out of it. We mostly ignored it, assuming it was nothing that concerned us, which ended up being our first mistake. Don't worry, we made more.

After using an improvised grappling hook Emmeral had fashioned to climb over the wall, we headed toward the church. The four dug up graves in the church's graveyard were our second ignored hint of what was to come.

Sometimes, we can be really dense.

We walked around the church until we say Brother Petrovich sleeping trhough one of the windows. A few hurried \textsc{Message}s were enough to convince him to let us in, though I nearly gave him a heart attack I'm sure.

Fortunately, his gods were friendly. We informed him about the werewolf bites and he led us through a prayer to his gods over the next couple hours. At the end of it, my \textsc{Detect Magic} showed a glowing, purple mass of transmutation magic swirling around my heart (but no one else's). Kinnaeia called on her gods to \textsc{Remove Curse} from me and we were good to go.

Kinnaeia didn't have the power to \textsc{Remove Curse} from all of us, if the priest hadn't been able to narrow down which of us were cursed\dots\ I don't know what I would do as a werewolf. Losing my mind to become a slobbering beast once every moon is not something I've ever wanted for myself.

Anyway, the priest seemed very worried about the graves being dug up. Remember those mistakes I mentioned? I told him the graves were why we were here and we would definitely look into the problem for him.

Yes, I lied.

Yes, that was a bad idea.

Heading out from there, we began with our mission: we sent Elwing (as a spider) to investigate the burgomaster's house, searching for Izek. He returned about an hour later with some news:
\begin{itemize}
\item the dungeon was completely empty
\item the sticks bundled around the house were gone, likely as kindling for the festival
\item Izek was in his room with countless dolls that were carved to look like Irena, sharpening his axe
\item oh and did I mention Irena was nowhere to be found?
\item the burgomaster and his wife were asleep though, as were the servants, so only Irena was missing
\item the broken mirror had been fixed or replaced
\item Viktor (at first we weren't sure since we hadn't yet seen his face, but we were right) was frantically searching for a book in the library
\end{itemize}

Emmeral suggested we try to get Viktor on our side and help us make a fool out of the burgomaster, but we all agreed he likely hated us too much. I mean, he certainly seemed angry when he \emph{nearly killed us twice}\dots

We left Emmeral to keep watch on Izek and headed toward the toymaker's shop, since we figured he must have gotten those toys from somewhere and this was the most likely place that we knew of. We were right.

Before we got to the shop, a large cloud of ravens seemed to be explicitly watching us. I need to remember to look into that. Maybe Strahd can see through the ravens' eyes?

The toymaker's shop had a sign on it: it opened at eight and would be closed midday for the festival. That gave us a much-needed hour to break in and prepare. A bit of slipperiness from our sneaky friend Emmeral and we were inside.

The shop was\dots\ pretty normal really, other than the Irena doll against one wall that was marked ``not for sale''.

With no hidden letter detailing exactly what was going on, we decided to interrogate the toymaker. And you know what? I honestly thought this would go well. We grabbed the toymaker (and his monkey, who Elwing has been obsessed with ever since), figured out from him that Izek was coming right at eight for the new doll (which he had the toymaker make a new one of every day until he got it ``right''), and set a trap.

Having convinced the toymaker that Izek was planning some secret scheme to hurt the burgomaster (I know, I thought it was quite brilliant), the toymaker agreed to let me impersonate him so we can lure in Izek with one condition: that we not let any damage come to the toymaker's shop, especially since Izek kept threatening to burn it down.

Yeah, I know. Just\dots\ not everything can go as planned, alright?

We sent Morianna to get Emmeral and set about planning: I used \textsc{Disguise Self} and put on a set of the toymaker's clothing, making myself look and sound like the toymaker, and we planned out exactly what would happen. I would have the doll on the counter, which I would stand behind, and I'd convince Izek to walk toward the center of the room. As he reached the point where all the toymaker's shelves meet, I would \textsc{Hold Person} him in place and the others would jump out from all sides. We didn't want to take any chances, the goal was to kill him as quickly as possible.

Honestly, the plan initially went off without a hitch. I lured him forward, he walked through the shop toward me, he didn't seem the to have any idea whatsoever that I was in disguise. We jumped out at him and hurt him really badly for he could really attack us.

With a giant magical battleaxe, I might add. I wish I could have recovered that\dots

Eventually, he cast a \textsc{Fireball} into the middle of the room. Yeah, as in the first floor of the shop we'd promised not to set on fire. Also, right into the middle of all of us.

The shop got set on fire.

In fact, the shop completely burned down.

We chased down Izek and had Emmeral kill him (he exploded, by the way. \emph{Exploded!} Why would that even happen?) while Elwing ran into the burning building to save the toymaker.

He's a brave one, I have to admit. But not the brightest; he couldn't figure out a way to save the too-fat-too-carry toymaker out of the building and he -- as Elwing tearfully recounted later -- died in Elwing's arms. He did save the monkey, though.

With Izek dead, we didn't have much time to catch our breaths. The toymaker's shop being set on fire attracted the attention of\dots\ well, of pretty much everyone. The others ran off to continue with the plan, while I stayed to continue the role of the toymaker and try to spread my story (that the burgomaster was in danger from folks like Izek that he'd stupidly put his trust in. Makes him soudn pretty incompetent, no?)

I convinced the people standing around that I was really the toymaker and that there was a plot against the burgomaster. That the brave and helpful adventurers were on the burgomaster's side and that all would be well. And, also, that they should help me put the fire out.

We tossed water onto the still burning building for a while, all while I continued to pass along the rumor. None of them seemed willing to listen too much, but I assumed at the time that was due to the giant fire in the middle of their village.

One of these days, I hope I'll learn to make better assumptions.

Eventually, some guards arrived and I passed the story to them; I explained that Izek had caused the fire, that the helpful adventurers had gone north (they'd gone southeast), and that they were working to help the burgomaster against a dastardly plot. They insisted that Izek couldn't have caused the damage, but I simply shrugged and insisted I could only share what I'd been told and seen with my own eyes.

They asked me to come with them, to speak with the burgomaster, but I asked I be allowed to put the fire out first (and slip away when they weren't looking\dots). Turns out they weren't really asking and went to put me in handcuffs.

I couldn't really allow that, given most of my magic only works when I use hand motions (note to self: figure out exactly what I can do with and without hand motions, so I can definitively know whether I can escape from handcuffs and such), so I \textsc{Misty Step}ped myself through an open window in the inn and hid in the dark room there for a while.

Eventually, I \textsc{Disguise Self}'d myself as one of the citizens I'd seen around town and walked out without any of them being the wiser.

It was pretty cool, if I do say so myself. I can see why Emmeral likes sneaking around so much, it's\dots\ thrilling\dots\ to completely fool people.

On my way out of the inn, I saw Rictavio leaving the inn and carrying a steak and followed him. He led the way to his caravan to feed the tiger, which -- surprisingly -- seemed completely tame. This may make our plan to release it a bit harder -- if it's tame, then it might not scare anyone. We might need to set it off somehow, make it angry.

That reminds me, I haven't written down our plan yet. I guess I'll get to it eventually, but the gist of it is to release the tiger during the festival and cause a bunch of other chaos, mostly to show the people that the burgomaster has no control over the city. The details were\dots\ less set in stone, but we were working on it.

I made my way back toward the toymaker's shop once it became clear that Rictavio was simply spending his time chatting with his tiger. I did consider speaking with him and trying to get him on our side, and in retrospect that would have been a great idea, but I thought he'd be unwilling to let us use his tiger as bait and ruin our plan.

Yes, yes, big mistake, I know.

Along the way back, I passed through city square, where I saw Morianna working alongside a villager, helping to set up a stage. She gestured toward an alley, where Kinnaeia was hiding.

Kinnaeia informed me of their efforts so far, few as they were. In the time I'd spent putting out a fire, spreading the rumor, escaping from guards, and tracking down Rictavio's tiger, they'd defaced a few signs and were mostly hiding around the stage in preparation for midday, when the festival would actually begin.

I sent a \textsc{Message} to Emmeral asking if he had any better ideas, since, mostly, Kinnaeia's plan was useless, but guards started patrolling the area and he had to hide under the stage. I sent a final ``I'll be back'' and went off to continue spreading the rumors. In the meantime, I later found out that Emmeral had put some damage into the stage to help cause chaos later. The first useful thing they did, I knew I could count on Emmeral.

Anyway, I spent a while walking around town, trying to convince the villagers of my story and spread the rumor. Gradually, I realized something was up: every single villager was completely uninterested in what I had to say at best and was terrified of hearing the rumors at worst. Eventually, though, I ran into some hunters and got a different conversation entirely.

They completely ignored my story and instead focused on me. ``Why do you talk so much?'' they asked, and I could see their thoughts churning. ``People from around here don't talk much.''

Which, you know, made me want to slap myself. Of course the people around here didn't want to hear any rumors, none of them were used to rumor-spreading at all since they \emph{never talk to each other}. Which is weird for a whole different reason, but at this point I wasn't going to go down that rabbit-hole.

I'm not going to say it this time. I know what I should have done instead, I'll leave it at that.

The hunters were convinced of my act as soon as I told them I was from Kresk and I headed back to Rictavio's wagon to chat with him -- I know what I'd thought earlier, but at this point I thought maybe he could help me share the story.

Along the way, I noticed Emmeral entering the coffin maker's shop and followed him in. The others updated me of their progress: Morianna had noticed a suspicious man entering a shop in town square with a large bag and leaving without it and had stolen some nails when she entered after him. She handed me some, which I threw into my bag with some confusion.

Emmeral had taken the hubcap off of Rictavio's wagon, then put it back on when he realized it needed to get all the way to town square before we wanted to release the tiger. Which, you know, should have been obvious. Maybe my faith in him is somewhat misplaced.

I shared my findings with the group, that no one in town was convinceable. To which Kinnaeia immediately suggested we kill them all. I cast \textsc{Detect Magic} to find out what mind-altering spell she was the victim of, but she came up clean. Something is seriously wrong with that dwarf and I'm terrified of when it will bite us in the back. No one else seemed to find her suggestion out-of-character.

We headed back to the town square around eleven, in time to watch the festivities. Eventually, the burgomaster appeared in a procession: his wife, son (we confirmed this soon), dogs, a bunch of kids dressed as flowers, and a bunch of villagers carrying a large wicker ball.

They entered town square, where the entire village was waiting, with all the pomp they could muster. The burgomaster walked onstage and turned to address the crowd\dots\ and promptly fell into the hole Emmeral had prepared.

Absolute silence for about five seconds, the entire town was quiet. Then, I had a brilliant idea and used \textsc{Prestidigitation} to make it sound like one member of the crowd had let out a single loud ``Ha!''.

More silence, this time shocked. Guards ran forward to lift him out of the hole. The burgomaster was fuming; so, of course, I used my magic to turn him the slightest shade of red and make steam billow from his ears. His wife tittered and Viktor smirked. The burgomaster shouted at the crowd to go back to their merriment and turned to his guards, which immediately split off in all directions, presumeably looking for me.

We ask Kinnaeia to give us a prediction as to the outcome of releasing the tiger now, since she imnformed us her gods would help guide us. Their response?

Releasing the tiger would be ``good and bad''.

\dots

This is why I'm not a cleric.

Soon, Rictavio's wagon showed up and the burgomaster began to get ready for what was clearly the main event. He lit a torch and began to wlak solemnly toward the wicker ball. So, of course, I \textsc{Prestidigitation}'ed the flames out. The burgomaster smiled thinly at the crowd and tried again, so I repeated my action.

This time, real laughter came from the crowd, from an actual villager, without me having to use my magic. My hero! I joined in with some (\textsc{Prestidigitized}) laughter of my own and the entire crowd soon started actually chuckling. The burgomaster begins ranting at the crowd, but the atmosphere was exactly right to deal the final blow to the crowds confidence in him.

I sent Elwing a \textsc{Message} to do his part in the plan, to release the tiger.

He sent back ``no''.

It would have worked as we wanted, the remaining gaurds would have been in place to subdue the tiger. No one would have been hurt, we just would have made a fool out of the burgomaster.

But Elwing said no.

At the last moment, he choked up and refused to do as we planned. And he's refused to explain his reasoning since then, insisting that the plan would have gone off badly even if he'd released the tiger on time.

That's when everything went to Hell.

The burgomaster stepped forward during his rant and saw Morianna out of the corner of his eye. Immediately, he sent the remaining guards toward us. All of them, every single one, including the ones we'd been counting on to \emph{keep the tiger from actually causing damage}.

I tripped the burgomaster on his way down the stairs and we bolted, all in different directoins. Most importantly, Emmeral ran toward the tiger's caravan to try to keep the plan intact.

He succeeded, I guess.

I mean, in the end, our end goal was\dots\ technically, we succeeeded, but\dots

The tiger lept from its cage and scared the crowd, who was already calling out about the burgomaster's incompetence. At first, we thought we'd succeeded as planned, that Lady Watcher could step in now and fix the problem.

Sometimes, we can be really, really, \emph{really} dense.

The tiger spotted the burgomaster's wife and something about her enticed it. It stalked toward her singlemindedly, pawing through the cloud without a single glance. When it got within range\dots

The tiger pounced at her and mauled her. It tore her to shreds, even with Viktor \textsc{Disintegrating} it almost immediately.

Rictavio tried to stop it, but he just couldn't. The tiger was too fast. No one was expecting it.

We sprinted away from there, heading toward the Watcher house as quickly as we could. The screaming of the crowd behind us only escalated, lightning flashed all around, the rising sound of undead groans\dots

We ran and we ran and we didn't look back.

\pause

At this point, I don't think any of us were thinking straight. We'd just killed a woman, not directly, I know, but she was innocent. And besides that, the screams behind us\dots\ we knew something terrible was happening and we were powerless to stop it. Labelas, we didn't even know what was really going on.

At this point, I don't think any of us were prepared for what came next.

We arrived back at the Watcher house and were greeted with open arms. ``Welcome, friends, I can tell all is well.'' is literally how she greeted us, with a smirk as she glanced at Elwing: ``not a spider this time, I see?''. I was too shocked to respond -- not shocked that she'd known, shocked at what had just happened -- but she simply led us to the main dining room, an incredibly extravegent room which simply oozed riches.

She snapped for a servant, which was just enough to snap me out of my daze.

She asked us how things went and I dutifully informed her of our success in terrifying the burgomaster, in killing Izek. She thought the entire affair marvelous and informed us that she never really wanted to be the burgomaster to begin with. That she was simply working for a better cause.

That was when Strahd stepped into the room.

\st{I still can't wrap my head around}

I felt confused when I saw Strahd, at first, until the enormity of what we'd done set in. We'd taken the city for Strahd, we'd done exactly what he'd wanted us to do, we'd\dots

He had been playing us, the whole time. And the worst part of that, is I'm still not entirely sure it was a bad thing. I mean, yes, clearly he's evil at this point, right? But that what we learned later\dots

Well, I don't want to get my thoughts out of order here. Just before Arabella's next prophecy, though\dots well, you'll see. That's when it started to make sense in my head. Not much sense, I'm still a bit out-of-sorts, but\dots again, we'll get there.

These people, the people in Barovia, or at least most of them, they're soul-less. They're nothing but constructs, they live and breathe sure, but not really. They don't really live, they don't have emotions, they don't have hopes and dreams and, and, and what does it even mean, to kill them? Are they really people? Is it more like killing an animal, or maybe it's even less than that? I don't know.

I really don't know.

\st{But if Strahd kills people like this, does that make him evil? It's the obvious thought when you see someone cause devastation on such a large scale, but if they're soul-less\dots I just}

Does it count? Does killing soul-less people count as murder? I'm not the most moral person in the world, I never strived to be some white knight hero, I fully admit that, but being the cause of so many deaths?

If they're people, then I can't stand what we did. If they were normal people with souls and lives and personalities I would be\dots\ I don't know what I'd be, but I wouldn't be able to continue as I have been, not really acting against Strahd, not really on his side. If they were people, my decision would be easy; I'd use this book to work against Strahd, to destroy him, to make him pay for what he's done. I \st{need}. I want to know what he knows, to learn all about the magic that courses through my veins, to walk the world for an eternity, learning more about my magic.

But to trade the deaths of hundreds, of more than that\dots\ I can't do that. I don't want to do that.

He'd played us. He'd used us as pawns, made us do exactly what he wanted. Maybe this was why we were in Barovia, to be his playthings, to give him exactly what he wants.

\st{I.}

We're going to work against him.

We're going to fight against Strahd. And I'm not giving up the hope that I can get everything I want out of him first, but that comes second. I've already learned so much from this adventure\dots\ there will be other ways to get what I want. This way\dots\ this is not alright.

Of course, with Strahd standing right in front of me, and without the knowledge that the people were soul-less, all I could do was fall back on talk.

``Thank you,'' he spoke grandly, without any thought that we were anything but perfectly in his grasp. We were, though, so really he was simply speaking accurately. ``the five of you are very effective. You've done me a great service today, giving me Vallaki, not to mention you've found my enemy for me.''

``Not at all,'' I responded, the sickness of speaking with him pushed down into my gut. ``We were happy to be of assistance. And what will you do now that Vallaki is yours?''

``Oh, but I have great plans! Wonderful plans! I can keep the zombies fed, I can spread my vampiric offspring all across the land\dots its truly a marvelous thing we've done. Why, you've even managed to break Viktor's ties to his mother -- the brat was always a bit too clingy for his own good. Simply wonderful.'' At this he looked at me shrewdly. ``Though this may not be the end of your\dots play-fighting with Viktor. I only have the time to take on a single apprentice, after all.''

Great to have that out in the open, where the others can see my temptation. Its been `absolutely marvelous' to have to deal with a Dwarf concerned I'm falling into darkness all evening.

At his comment, I smirked and tipped my head, ever playing the showman for him. Maybe that's why Kinnaeia is so convinced I'll be trying to go kill Viktor? Not that I wouldn't mind gaining access to his spellbook\dots or getting some payback for him nearly killing me twice\dots or having Strahd absolutely convinced I was his underling and willing to teach me all of his arcane talents\dots

Alright, I can understand why she's so worried about me.

And maybe I can get all that I want out of him, but killing Viktor in cold blood to become Strahd's apprentice and simply pretend all of this is alright? I can't do that.

``As a reward, you may have the Petrovich house. Viktor's lab, of course, has already been moved to Castle Ravenloft. While we're on that topic, I'd like to invite you to my wedding in four days time. No gifts are required, of course, simply your presence in watching me wed the lovely Irena.''

Yeah, that's where Irena is.

With that, Strahd left us -- presumably to prepare for his upcoming nuptials. For a few moments, each of us were stuck in a trance, processing what we'd just heard. Lady Watcher snapped out of it first, to congratulate us once again and lead us out. As a parting question, I asked her who would lead the guards now, hoping for a few potential new allies. Her response was less than helpful.

``Don't worry, they'll be dealt with.''

I am, in fact, worried about it.

We wandered aimlessly for a time, only coming back to our senses as we watched a zombie tearing chunks from a fresh corpse. I'm not sure who said it first, but for once we were in complete agreement with each other: ``we should find Rictavio''. Maybe it was that he was our only real ally in the city, maybe it was because we felt bad about using his pet as the engine of the destruction of the city. Maybe we just needed something straightforward to focus on.

Whatever our reasons, we were set on our goal. As one we raced back toward town square, killing a few scraggling zombies along the way and avoiding the larger hordes. I hardly remember stopping, only running past them and killing with barely a thought, my magic flowing around me.

Eventually, Elwing thinks to \textsc{Locate Monkey} and we re-oreinted back toward the church. We find no monkey there, but we do find barricaded doors and windows. A knock from Morianna terrifies the people inside, but Brother Petrovich opened the door a crack to speak with us. He informed us that they were barricading and praying. We wanted to take them with us, we tried our best to be convincing, but they refused -- the church was pretty much the safest place in this entire Labelas-forsaken country. We couldn't really argue with that one, especially now that they had the protective bones back once more.

I hope they're alright in there.

Eventually, we decided to leave them behind and follow the monkey tracks that Emmeral had discovered during our talk. They led almost directly to the -- now undefended -- West Gate of Vallaki\dots more specifically, they led directly to a zombie standing in the middle of the path, leaning on a cane, and wiping dirt from its jacket.

We paused for all of a second before Morianna burst out a greeting: ``Hey Rictavio!''

He turned to face us in fear, brandishing his cane to reveal a hidden sword. Our conversation with him was not very useful; it pretty much went like this:

``What have you done to this city?!?''

``Yeah, sorry about that. We didn't mean it.''

``Well now I have to hide from my enemy.''

``Yeah, Strahd knows you're here.''

``No kidding. Thanks for that, by the way.''

``Our bad. Also he's marrying Irena.''

``Yeah, that makes sense. She's the reincarnation of his long lost love, so that was pretty much bound to happen.''

``Wow, we did not know that. Can we help you out at all?''

``No, you've done enough. You've ruined everything and I've lost every advantage I had. Now I need to disappear.''

``Wait, does this mean you know how to leave Barovia?''

``No, of course not. But I do have some places within Barovia that I can go.'' he sighed a bit here and pushed Piccolo toward us. ``You there, druid, take Piccolo, he won't be safe with me.''

``Can we help you fight Strahd?''

``Fight Strahd? I'm not fighting him, I'm running from him! If he finds me, I'm dead. Look, I need to leave. I might be in touch if I can manage it safely. Try not to make things even worse while I'm gone.''

I may have summarized what he said a bit, but the tone was there. Normally I'd take offence to being spoken to so rudely, but\dots well, I can't say we don't deserve it.

With that, he disappeared. So much for that ally.

Speaking of allies, though, we decided to head toward the Vallaki camp.

We headed out to the Vistani camp, hoping that they hadn't been attacked by the undead army. They'd been more than friendly to us, maybe we could even convince them to take up arms against Strahd! It would be nice to have someone on our side, for once.

\subsection*{The Vallaki Camp's Party}
The Vistani were partying, celebrating ``our success in Vallaki''.

``Lovely, that was our goal,'' I sighed back. I'm getting tired of having to lie to pretty much everyone we come across, just so they don't know they're our enemies.

Apparently the Vistani have always been supporters of Strahd, we even got a history lesson out of the deal: apparently, years and years ago Strahd had led an army to capture this country. He had fallen in the fight, but the Vistani had saved him, earning them pretty much eternal gratefulness. Now that he's all-powerful, the Vistani are more than happy to continue serving him in exchange for special favours.

Like protection from all of Strahd's evil, apparently the Vistani are completely safe from all of Strahd's schemes, from all of the zombies, from vapires, from ghouls, from pretty much everything terrible about this country.

Lucky kuu'dattora.

As Kinnaiea began a drinking competition once again to help loosen their tongues, we asked a few more questions: we finally had a chance to ask about ` a vampire named Arasimus', but they had no idea what we meant. They directed us to `Madam Ava' -- the oldest living Vistani -- in the Barovian Vistani camp when we requested to speak with the wisest person in the country. They informed us that they happily worked for Strahd by running expiditions beyond the borders -- though they could only leave with his special permission -- and taking care of ``distasteful'' elements of society at Lady Watcher's request.

The whole thing left me feeling a bit slimy, but at least we got some horses out of the deal. Apparently Strahd had lavished them with gifts in celebration of our victory and they had more than enough horses to spare one for each of us. It'll help us run away in fear somewhat faster, at least.

As Kinnaiea drank, Morianna stayed with her and Elwing went off on his own. Emmeral and I planned our next moves -- at this point, I think we understand each other pretty well. Neither of us was alright with what had happened; we needed to know how to put an end to this. We agreed to set out to the Wizards of Wine winery at first chance, searching for the knowledge promised to us in Arabelle's prophecy.

In her first prophecy, that is. I told you I'd get to this, didn't I?

We had just spotted Elwing having a conversation with a Dusk Elf and were about to join him when Arabella peaked her head out from behind a wagon and gestured us over to her.

``What do you know about Madam Eva?'' she asked us, getting immediately to the point. ``It doesn't matter -- look, Strahd brought you here with some plan. He made you meet the wolf-touched. He knows your actions before even you do. How, you ask? Madam Eva tells him. She's a psychic, a prophet like me, and she's very much on Strahd's side. I hate him, though, since he keeps the soul-less trapped here.''

``We're in a demi-plane, trapped in a pocket dimension of Strahd's domain. Most of the villagers here are hardly even real people, since there aren't enough souls to go around. I just want to see him defeated so those soulds can move on.''

``There's a book you need to seek out, at the Wizards of Wine winery. So long as you hold that little piece of Strahd's past, he can't have Madam Eva see your actions -- he can't have anyone scry your actions. You need to get that first, to have any chance of maneuvering without his knowledge.''

Arabella blinked a few times after that, screwed up her eyes in concentration, and then burst into a trance to deliver a second prophecy: ``the wolf-touched has gone to resolve her past''. She recovered pretty easily from this one -- and even explained that she'd done it on purpose.

``True prophecies will confuse Madam Eva's sight. You have several hours where she will be completely unable to predict you, I suggest you make the most of the time. Strahd will notice the hole in her predictions, though. He'll know I'm involved. I need to leave this place, but I have faith in you.''

Isn't that nice, she has faith in us. Despite the fact that we'd found and lost our second ally in as many hours, I couldn't help  but feel somewhat optimistic about the whole thing. We're trapped in a demi-plane, under control by a vampire that wants nothing more than a world full of suffering\dots but we can fight back. With enough power, we can defeat him and unravel this entire plane. We can save\dots everyone. That's a pretty great thought.

Except for the part about how the ``wolf-touched has gone to resolve her past''. It took us a bit of searching, but eventually we realized: Kinnaeia and Morianna had slipped away with their horses.

They'd better hope they left for a damn good reason.

We met up with Elwing and the three of us tried to decide what to do; I say try because we could hardly make up our minds. We needed to go after the book to remain hidden from his powers, but without the girls we would have an even more difficult time if there were any obtacles in our way at the winery -- and we're pretty sure there will be, since that seems to be how this country works. According to Elwing, every Dusk Elf hates Strahd and they both know where his power source is and are willing to bring us to it -- we seriously considered beelining it straight there, but we're pretty sure the defences there would overwhelm us. Better to be invisible to Strahd, first.

After far too long spent arguing, we decided to head first for Morianna and Kinnaeia and then head to the winery immediately thereafter. We grabbed some horses and a wagon from the Vistani and set off, only pausing to let Elwing inform his new Elven friend (we have an ally that hasn't left us yet! I'm so happy) that we'd return soon.

As we head toward the wolves' den, I can't help but wonder: did we make the right decision, going blindly after a Dwarf and a Human that had just abandoned us without a word rather than take our chances with a smaller group at the winery?

\pause

Well, I'm not sure we can trust them not to run off again, but they seem happy to help us find Strahd's book, at least.

We travelled toward the general area we knew the wolf den to be; at first, we thought we'd have to search carefully to find it, but it turns out they marked their cave with a giant stone wolf head. Subtle, these werewolves.

As we approached the cave, we began to make out a few figures: Kinnaeia and Morianna outside the cave and an old man and a young woman within the cave, surrounded by wolves. The woman darted toward Morianna, as if to attack, but stopped a foot away. We couldn't make out most of the conversation from how far away we stood, but as far as I understand it the woman was the new alpha to the werewolves now that we'd, uh, gotten rid of the previous one.

Morianna asked her why the werewolves ``attacked that house that night'', which I assume means I missed some context there, and was told that ``he told them to kill everyone, but spare the girl''. And that now ``he has something precious of mine, so we can't act against Strahd''. That was my first clue as to the context, by the way. Sounds like Strahd sent the werewolves to attack Morianna's family? I wonder why he would so that.

Morianna offered our help in retreiving whatever precious thing the alpha was referring to in exchange for the werewolves' help in fighting Strahd. The alpha seemed to have tentatively agreed, but only if we can rescue her mate ``Amil'' from Strahd's dungeon.

Yeah, we'll have to break into Strahd's castle once just to get help breaking into his castle.

I hate this place.

Anyway, the werewolf (who Morianna referred to as ``Salika'') gave Morianna a whistle to identify us to Amil and sent us on our way. We ushered them back to our wagon, told them about Arabelle's latest prophecy, and are now riding onward to the Wizard of Wines winery. Here's hoping this goes smoothly.

% \pause

% \section*{I Hate Druids}
% It didn't go smoothly.

% I mean, sure, we managed to get the book, but the hundreds of tree-things in our way\dots well, we're all pretty glad Elwing can think on his feet.

% When we approached the winery, a


%%
% find winery, guy in black cloak (damian martikov) beckons
% we approach: "here to save my vineyard?" "frankly no, but we might be able to help you. ours is time-sensitive, have you seen a book?"
% I am damian martokov owner. for past couple weeks, winery under siege by evil druids and walking treefolk.
% We have druid that may be able to help
% Me and family driven out, we'd be greatful. tell prophecy. hmmm, reminds me of vats of sand for glass blowing within wooden barrel.
% barrels are in southern room of winery
% saw 3 druids, dozen trees, why attack? maybe to steal awesome winery. we're very land-respectful
% to winery (leave wagon):
% tree things start to attack
% call out in sylvanic
% dan calls out after "speak with plants"
% "must obey the galthus"
% ginelle: galthus tree is tree grown around vampire that spawns blights
% mage hand open back door, rush in
% start looking around, trees rush in from downstairs
% ginelle kills w/ radiant flame, I kill w/ firebolt
% james slices one to death, i firebolt another
% door explodes into james, reveal druid and bunch of tree kids
% thunder blast, tree kids run in. ginelle throuws oil, I cast firebolt. FWOOSH!
% james runs after druid, he collapses and bleeds profusely
% other druid screeches in druidic about them obeying because vampire staff
% lion-dan jumps over fire and pounces as druid, vine wall comes up at old door and blocks him off
% big treethings from outside break in through windows
% james kills druid, we run downstairs. i prestidigitation a fire trap
% a couple jump down, but many scared
% black cloud comes down and kills them all
% dan tells us of druid with magical stick controlling them, he took it and broke it
% we search sand and stuff, find book (w/ unknown crest). mostly unreadable, but:
% HOLY SHIT I'M STRAHD I USED TO BE COOL BUT HATED MY BROTHER AND SHIT AND BECAME A VAMPIRE FOR TATIANA
%
% dan gives me remains of vampire staff
% ginelle detect magic; notices magic wine (evocation), strahd book (abjuration); I identify as poison potion
% Me and Lara take poisoned wine bottles
% Explore basement, ginelle notices secret door
% detect magic, see nothing
% lara opens door, see 15ft tunnel with mold. dancing lights, see small room
% cast ray of frost, shrivels. shit ton more, pacify the mold
% cast firebolts to regrow mold behind (fridge)
% explore upstairs
% try to identify mirror in master bedroom, nothing. james looks at chest, lara finds deliveries to strahd in desk
% find kitchen, eat jerky
% "bucephilus" (famous demon nightmare horse) written on toy
% printing press with labels, label poisoned wine as Red Dragon Crush (mislabelled, color of best wine). desk locked
% head outside, more people
% ginelle: "we killed druids, stuff posioned but I can fix it", "thanks, how repay?" "sleep here?"
% "stalled on propduction, three magic gemstones produced wine (invigorate) but blights stole gems". went southeast
% "vistani asked for wine, can we haz?" "yup! at least three"
% let's rest, tell you if we'll accompany caravan in morning
% dan casts mending on staff at my request
%

% decide to follow after stolen things to get the druids

% \sleep
