% The Twelfth Day of the Ninth Moon, Year 735

\section*{The Watchers}
When we woke up this morning, Emmeral shared his knowledge of the dragon in the haunted house with the others. He seems very excited about the possibility of a sun sword; seeing him eviscerate our enemies makes me agree pretty strongly. A sword of prophecy? That's exactly what I want our best warrior to carry.

Emmeral argued strongly for heading toward the dragon, Morianna argued for heading to the Wizards of Wine winery near Kresk. We ended up compromising; I convinced them to talk with Lady Watcher before leaving the area -- something about what we'd heard of her set me on edge, I didn't want to get on her bad side. Given the whole demon-thing, I think I made the right decision.

Our argument got a bit too loud, loud enough to wake up a groggy Luvash. He was initially angry at us waking him, but we calmed him down by offering to pick up some more wine for them from the Winery -- we knew we'd head there eventually. He offered us a caravan to carry the wine, so we promised we'd return soon to fetch them. In the mean time, we headed toward Vallaki.

I still can't believe we went back there. And that we're going back again tomorrow. Terrible city that it is\dots\ but hopefully tomorrow will fix that.

We should have tried to sneak into the city, none of us had realized we'd been banned. The guards immediately moved to close the gates against us, but a handy bit of magic saw us through: I cast \textsc{Invisibility} on myself and Emmeral, then \textsc{Misty Step}ped through the gates right in front of the guards. Elwing became a horse and galloped in behind us; the others were trapped outside, being unable to cover that much distance without being stopped by the guards.

And so we were split; Emmeral, Elwing, and I met around the Watcher family house, and Morianna and Kinaaeia explored in the area around the city.

Being invisible and trying to meet up with an invisible ally was\dots\ interesting. We had a vague understanding that we would both be heading toward the Watcher house, but heading there together wasn't really an option. We also had to rely on our memories of the location of the house, since, you know, invisible maps.

When I approached the house, I noticed the family's scarecrow had been moved to point very clearly toward their back door. Note to self: Emmeral is not quite as subtle as he potentially could be.

I turned myself visible again while continuing to concentrate on Emmeral's invisibility and knocked on the rear door. Emmeral acted wonderfully here; without needing to discuss any plans, he acted perfectly as backup for the entire talk with Lady Watcher.

Upon knocking on the back door, a servant answered. He asked why we were here, I informed him Lady Fiona requested our presence. He asked why we were at the back door, I insisted the ``burgomaster has eyes everywhere''. That was enough for him; if tomorrow does not work out in our favour, I'll need to remember this.

After a moment, the door swung open.

The Watcher family's house was arranged along a path, doors left strategically opened and closed to lead us directly to the dining room. I very carefully followed this path -- I wanted to make sure I put the right foot forward. In the dining room, I found Lady Watcher sitting alone.

Well, I say alone, but she wasn't -- not really. A small red demon -- some sort of imp, perhaps? -- sat on her shoulder, watching me for the entire conversation. I didn't react to it, and nor did it speak, not that I'm even sure it has that capability.

Is it evil to be working for a devil? Or, at least, is it too evil to justify? Clearly its not a good thing, but\dots\ its worth it, right? We've been banned from this city, the biggest city of Barovia, where we might be trapped for who knows how long. And the city is so clearly rotten to the core: there's vampires, necromantic mages, half-insane burgomasters\dots\ and Lady Fiona offered us so much in exchange for helping her.

No, we're doing the right thing here. Helping ourselves well also making the city better for all of its inhabitants -- what better action could we possibly take?

Speaking with the Lady Watcher was an exercise in charisma. The entire conversation I felt like we were dancing (or sparring, I'm not quite sure\dots). We jousted back and forth on the specifics, on whether it was obtainable, but really, it comes down to this:

In exchange for assassinating Izek and embarassing the burgomaster before removing him from his position for that the title will pass along to Lady Watcher rather than his son, Lady Watcher will gift us with property in the city, her general aid and friendship, gold, and other rewards.

I didn't quite accept this mission on behalf of the others, but I did heavily hint that this should be acceptable, pending speaking with them. And now that we've convinced the others to go ahead with this in the morning\dots\ well, we can't really back out now.

After leaving the Watcher house, we made some plans: not very good ones, apparently, since we went back on them almost immediately, but I guess that's how it goes. We sent Emmeral to hide near one of the city gates to help us sneak in over night and went ourselves to the general store to find some disguises for the others.

Ah, I forgot to mention: Elwing had caught up with us, in spider form, somewhere outside the Watcher house, and had listened in on the entire conversation.

I spent far too much gold on hunting outfits, like we'd seen here and there around town, then we headed toward Emmeral's likely hiding spots. Since we had the disguises, there wasn't really any reason to stay hidden within the city.

We left through the eastern gate and looked around for the others: they weren't there.

We headed to the gate we'd entered through that morning: they weren't there.

We headed to the Vistani camp and chatter with Arrigal: they weren't there and Arrigal hadn't heard anything about them passing through after we'd left.

With no idea where they were, we thought about where they'd be likely to go. Since Morianna had been so insistent on heading to the Wizard of Wines winery this morning, we assumed they'd head that direction and so followed along the road toward Kresk.

\section*{The Tower of Khazan}
Eventually, we noticed some muddy footprints heading off of the road and found them walking slowly toward Lake Baratok. It turns out they, too, had an exciting day; they'd found a bag of clothes off the road that they were confident had been left by a werewolf, perhaps even the werewolf our prophecy had mentioned. They'd left her(?) a note saying to meet at a bridge on the Raven River, near the winery. She'd forgotten to set a timeline for this, so we fixed that; after some discussion we argreed to give ourselves five days to deal with tomorrow's festival, any fallout from that, and anything else unexpected that comes up until then.

The reminds me: today is the ``Twelfth Day of the Ninth Moon of the Year 735", as the Barovians keep their calendars. Our plan is to meet on the bridge on the seventeenth.

A few things that worry me about Morianna: as she was filling me in on what they'd been up to, she informed me they'd found a bunch of herbs and set them on fire, then she'd sat and whittled a doll for Arabella. Upon me telling her about our plan to embarass the burgomaster, assassinate Izek, and help Lady Fiona take control of the city, her only response had been to suggest we set his house on fire.

What is wrong with this woman?

Emmeral had gone to investigate the clothes bag and confirm the werewolf tracks were, indeed, made by a werewolf. Upon his successful confirmation, we decided to head toward Lake Baratok and take a look around before dealing with the Vallaki festival tomorrow. Kinnaeia also made a poison out of hemlock to kill the lice we'd gained from our makeshift disguises; we kept her spare dosage and hope it might come in useful tomorrow. The fact that our otherwise unassuming cleric knew to make a poison out of hemlock\dots\ should I be worried?

Either way, Elwing and Morianna set off toward the bridge to leave her mark somewhere the werewolf would find it and the rest of us headed off to meet them at the lake. By this point it was about mid-day; we had the afternoon and early evening to explore.

Kinnaeia, Emmeral, and I, while walking through the forest, came across a ``friendly'' traveller (Kaltrin the now-dead werewolf) who warned us of the danger of wolves in this area. He also warned us that werewolves haunted the area.

No kidding.

He gave us an amulet to protect us from these dangers and set us on our way; I cast \textsc{Detect Magic} to determine what sort of protection it would offer, but found it was unenchanted. Casting \textsc{Identify} to learn even more turned up nothing interesting, only that it'd been made by hand. Like fools, we wished him well and carried on, only for him to ambush us with a few other werewolves on our way out of Van Rikten's tower.

We met up with the others near the lake. Immediately, Morianna insisted the pendant's symbol is the one she's been searching for and asked for the pendant. She's tight-lipped about why she's looking for this symbol, or what she knows, but my best guess is that this is the symbol of the local werewolves. Why would she be looking for a werewolf symbol, though?

Once we arrived at the lake, we found a decrepit tower built on an island in its centre. The island was clearly built specially for the tower, maybe by magic? It was impressive, to say the least. The tower itself? Less so. It was falling to pieces\dots\ the golems, on the other hand, were interesting: they responded to thought, controlling the elevator instantly.

Now that would be cool! I wonder how whoever built this tower created the golems? Having a golem or two to protect me\dots\ hmmm.

There was originally a Vistani wagon sitting outside the tower but we, uh, broke it. Emmeral picked the lock, and I \textsc{Mage Hand}ed the door open. So, of course, given our luck, the wagon instantly exploded in a giant fireball that echoed across Barovia.

Yeah, we're bad at this.

Speaking of being bad at this, the door to the tower had an emblem with several stick-figure drawings arrayed around it and the name \emph{Khazan} written above it. So, us being us, Kinnaeia threw a rock at it then ran forward to touch the door, which immediately exploded, catching us all in a field of lightning. Idiot.

Besides the trapped front door, the tower also had a dome of anti-magic. It seemed to both disable magical effects and prevent casting them within it\dots\ now that was terrifying. Being unable to cast spells? I could hardly stand it, I was terrified the entire time we were inside.

Given the front door was trying to kill us, we decided to walk around to find a different way in; some half-broken wood scaffolding leading up to a window worked for that. Well, sort of worked: Morianna climbed in the window on the third floor and we immediately heard two thumps as she fell through both floors to the main floor. Again, we're not very good at this.

Either way, she opened the door for us and we explored the tower. This is where we found the golems; four of them standing around an elevator platform in the centre of the tower. The elevator was controlled by our very thoughts; now \emph{that's} something incredible -- can you imagine being able to pre-cast spells and let them respond to my thoughts? Imagine preparing an area with hundreds of \textsc{Firebolt}s, all ready to fire at my command! So long as I had some time to prepare, I'd be ready for anything!

I wonder if there's anywhere I could find someone who knows how to do this? Maybe this Van Rikten person can teach me.

Anyway, Emmeral, Morianna, and I headed up to the top of the tower, which was furnished. It either belonged to Strahd, or Van Rikten, or maybe Khazan is a person, or\dots\ I don't know, I don't think we can be sure. There was a portrait on the wall of Strahd and a younger male that appeared to be him, who we think might be Van Rikten, unless Van Rikten is an alias for Strahd, but\dots gah! I can hardly keep it straight.

Alright, one at a time:
\begin{itemize}
\item we found a suit of armor that wouldn't budge, which -- given that tower -- was probably because it was enchanted to attack us on command or somthing like that.
\item a chest in the corner contained a dead Vistani head and a lavender-scented bottle. Likely, the Vistani head belonged to the Vistani mentioned in the memoirs. The lavender solution (which I took, just in case) appeared to be for embalming.
\item a few pages of Van Rikten's journal, which Emmeral read to us. Let's see what I remember of that: Three decades of investigating to expose creatures of darkness. Van Rikten was a hero, sage, and master hunter. He survived supernatural assaults and was obsessive about a vampire that killed his kid, so he became bleak over the ``weight of time''. The man lived too long and either was, or was like, a lich bound to existence, to endure forever. He wanted to die, but didn't think he will. He thought we wouldn't be considered a hero for long, since he was alright with being the indirect cause of many deaths. He was the target of a Vistani curse but hadn't born the brunt of it, rather his friends were. His child (Arasimus) was taken by the Vistani and sold to vampires; Van Rikten stalked the vampire down and killed him, then tracked down the Vistani and tortured them for information. The Vistani didn't kidnap his child in secret, mind, but Van Rikten couldn't save a sick Vistani that he promised he could save and offered them ``anything he owns'' to make them feel better. After an hour, he realized they took his son. On his horse, he took the dead Vistani and followed. When he decided to avoid camping at night, he was set upon by the undead -- a lich intervened to spare him, though. Lich detected him, took control of the zombies, warded him against undead, animated a dead Vistani to lead to camp. He found the Vistani camp, the dead still following him, threatened the zombies upon the Vistani, but they said they sold the son to Baron Mateuse. He released the zombies to kill the group. The leader of the Vistani cursed him to ``live you always among monsters and see everyone you love die under their claws''. He found Arasimus later, but as a vampire who wanted to die, so he killed his son. He thought the curse was done then, but it wasn't and he got angry.
\item a map of the area with a bunch of places marked off. Mostly places we already had heard of, but getting an exact location was exciting.
\item a portrait of Strahd and a younger man that looked very similar to him.
\end{itemize}

So, either Van Rikten is Strahd, way in his younger days, before he actually became a vampire, or Van Rikten is Strahd's\dots\ younger brother, maybe? Either way, somehow they are definitely related. As for Khazan\dots\ maybe it's Strahd/Van Rikten's younger brother's name? Another alias? Someone else, who built the tower before Van Rikten took it over? Not even a name at all?

Kinnaeia found ``Van Rikten's Guide to Vampires'' in the wreckage of the wagon outside, which definitely makes me belive Van Rikten was a vampire hunter at some point. If his journal isn't accurate, he at least put in the effort to make it seem like it is.

I've attempted to say \emph{Khazan} as a spell, with magic swirling inside me, but nothing happened. Maybe it's not a spell, or it is, but it only works near the tower? It may be worth going back there at some point, it could be a safe haven when we need one.

On our way out of the tower, we were intercepted by werewolves who wanted to kill us and steal our supplies, one of them being the guy we'd seen along the road earlier.

Now at this point, I put to use the knowledge of magic that I gained last night to amazing effect. I was definitely right, reading through this spellbook has increased my magic by leaps and bounds.

Immediately, I let the magic flowing through me burst out, casting two \textsc{Firebolt}s at once -- one from each hand. The closest two wolves positively exploded in flames, my magic incinerating them to the core. This burst of magic, though\dots\ I have no idea how, but the others told me later that I turned into a potted plant.

A potted plant.

I just\dots\ I just. What?

I have no words.

When I\dots\ came back to myself\dots\ Morianna was laying swordless near me and Kinnaeia looked half-dead\dots\ basically, without my magic they got completely demolished.

My magic was particularly active after that, I could barely believe what I could do. Immediately, I spun toward the three werewolves and cast a \textsc{Firebolt} at one with one hand and \textsc{Hold Person} on the other two with the other. Holding two of our opponents still with only a single hand, the others dove in and killed most of the small wolves. The third one, though, they completely ignored until it dove on me and bit at my arm.

The bite was enough to make me lose my concentration, allowing the other werewolves to finally start attacking us.

Morianna knocked him away after he bit me, but the damage was done: the other werewolves dove at Elwing and Emmeral. I cast \textsc{Magic Weapon} on both of Emmeral's swords to give him an edge against the werewolf that was slashing at him and \textsc{Fire Bolt}ed the one that bit me, all within seconds. The amount of magic I've been able to output lately and so quickly\dots

Emmeral, in usual fashion, tore through the werewolves once he had a second to catch his breath. Elwing, too, made a great showing; somehow he'd learned to \textsc{Call Lightning} to strike at our enemies\dots\ pretty impressive. Not quite as impressive as my command on magic, lately, but\dots\ well, we can't all understand the intricacies of magic quite like I've been learning to.

Well, sort of.

The fight eventually ended -- after I had another couple chances to demolish the werewolves with a flurry of spells -- with the lead werewolf running away in fear after we'd knocked down his last surviving companion. Emmeral charged after him, killing him before he could escape, and we turned to the wolf that we'd knocked down.

I tried to intimidate him into talking, infusing my voice with my magic, and somehow ended up with a third eye on my head.

At this point, I just\dots\ it's not that I'm losing control of my magic -- far from it, I have more power now than ever before -- but\dots\ well, I need to spend more time meditating about this.

We tied up the wolf and spent some time calming ourselves down after the fight and having Kinnaeia patch us up. I sat for a while to think about how my magic was changing, but really didn't get anywhere with it. After I finish updating this journal, I'll spend some more time\dots\ I definitely need to get this more under control before anything bad happens.

\pause

That reminds me; Kinnaeia ate an entire wolf while we were sitting around and collecting ourselves. I have no idea how to react to that.

\pause

After we'd rested up a bit, we started interrogating our prisoner.
\begin{description}
\item[Why did the wolves attack us?] It looked like we had equipment to steal.
\item[What does the symbol on the necklace mean?] Nothing, it's just their clan symbol. Morianna was insistent on asking this, though she refused to tell us why, and seemed confused and disappointed at this answer. That woman\dots\ she only gets more and more suspicious as time goes on.
\item[What happens now?] The beta will be in charge, since we just killed the alpha (oops).
\item[Where is their den?] They have a cave across the lake to the west.
\item[What's in the den?] A bunch of children that they're planning on turning into werewolves at some point.
\item[How does one become a werewolf?] Get bitten and wait about a day. As far as the werewolf knew, there's no way to prevent it, though maybe a priest would know more.
\item[How many packs are around here?] Only the one, which covers the entire area west of Vallaki.
\item[Are the wolves transformed werewolves?] No, just pets.
\item[Do the werewolves like Strahd?] Yeah, they're generally a fan. At this point, Kinnaeia butted in to ask specifically whether the werewolf ``knows anyone of them that doesn't like him, maybe a woman?''\dots\ talk about making it blatantly obvious. Either way, the werewolf pretended to have no idea what she was talking about though eventually folded and told us the new beta (``Zallika'') was always against both the former-alphaand Strahd.
\item[Whose clothes were in the bag?] Could be anyone's.
\end{description}

At this point, we pretty much started yelling at each other about what to do next and why. The werewolf took the opportunity to stark slinking away, so Emmeral stabbed him. This, of course, caused even more arguing\dots\ I swear we spent nearly an hour yelling back and forth.

It's a nightmare, really, its like we have no common goals but somehow we're working together\dots\ well, sort of. I have no idea why we're even staying together\dots\ I guess we're all pretty confident we'd die on our own in this Labelas-forsaken place. In retrospect, that's a pretty good reason to stick together.

Eventually, we settled on a course of action: Emmeral skinned the alpha's pelt in case we needed future proof, we threw the rest of the wolves into the lake, we headed back to the Vistani camp, and we planned to get the priest to de-curse us in Vallaki the next day.

We're turning in early so we'll be well-rested for our city invasion over night. The plan is to head out around two or three in the morning. I'm so restless I'm not sure I'll be able to properly trance tonight\dots\ I just hope our plans work as well as we're hoping.

\sleep
