\section*{How My Story Really Begins}
After leaving the Fae, I wandered through the forest a bit more. My rough plan was to return to Neverwinter to figure out how long I'd been with the Fae and to get some more supplies before setting out to\dots\ somewhere. I certainly didn't have a location in mind.

Along the way, I ran into Scathac.

Scathac was an interesting elf. She called herself a monk, but I'd never heard of a monk anything like her. She had a massive hero complex, and she was anything but calm and sedate (you know, the way monks should be). She was always willing to rush into danger without thought for her personal safety\dots\ which was convenient, given I met her as I was being set upon by bandits.

The bandit thing was completely my fault: I'd been wandering with my head in the clouds, thinking of future plans and not looking where I was going. I've learned to be a bit better with that, given this country is clearly out to kill me, but back then (still only five days ago? I can hardly believe that) I had some problems.

Anyway, these bandits jumped out at me as I walked along a small, twisting path through the woods. They noticed the crystal in my hand and the full pack on my back, I assume, since they demanded I hand them over. I'd only barely shaken myself enough to focus on the present when a blur ran out from the trees to punch one of the bandits straight in the face.

I remember looking at the bandits, them looking back at me, all of us with confusion on our faces. Scathac yelled, breaking the confusion, and the bandits scattered.

Turns out, Scathac had been apprenticing under some recluse somewhere, since she was ``destined to be a hero''. She was clearly delusional, but I can't deny her hero complex was useful in saving me, not to mention in helping us survive the murder house. \dots\ I'll get to that later.

We decided to travel together for a while -- her, searching for adventure, me, searching for magic. We continued heading toward Neverwinter, since it was pretty easy to convince her that adventure lay that way. Having someone watching my back\dots\ it was nice.

That day was clearly one for meeting new people; we ran into two other travellers within hardly a few hours: a druid who seemed intent on bringing what I was mostly sure was a dead body to the nearest city and a ranger who didn't seem to have a real reason for being in the woods in the first place, but seemed content enough to be our eyes and ears. I was happy anough with the arrangement: I no longer needed to distract myself watching for bandits and could instead focus on working out my upcoming plans.

Elwing, the druid, is an odd character. He seems to be motivated mostly to find animals\dots\ I have no idea why, but he often asks people where he might be able to find some new animal he'd never seen before. Eh, druids, what can I say? He seems like a pretty ``good'' elfling, and he can't lie to save his life (as well as I can tell, which is pretty well), so it's pretty easy to figure out how he thinks. He gets a bit distracted by ``doing the right thing'', but so far me and Emmeral have been able to convince him to avoid that (for the most part).

What do I say about Emmeral? He's a ranger, and a damn good one at that. I've seen him slice through multiple monsters in a single swing and be at the front lines taking down groups of our enemies. He can't really take a hit, but he's pretty great at killing before that becomes a problem. I would not want to be on his bad side, that's for sure. We've been scheming together lately: he wants to leave our group and I'd be inclined to go with him. The fact that he's talking with me makes me pretty confident I'll be on the right side if he ever does decide to go through with that.

Anyway, the four of us were travelling along for a while when we came across a couple dead horses and a broken down wagon. Once Emmeral nudged me and I noticed the scene, I figured it had been picked clean by the bandits that came across us earlier.

Turns out that assumption was incredibly wrong.

Emmeral went ahead and investigated the scene a bit -- and that's about were our adventure began.

Turns out there were Goblins about, Goblins who were still watching the scene in case any dumb fools like us came across it. They ambushed us and tried to kill us, as Goblins do. Emmeral though\dots\ Emmeral was terrifying. He killed almost all of them, though one got away with a missing ear. I'm proud to say I helped: I threw a bunch of \textsc{Fire Bolts} at them, enough to hurt at least two of them.

This is where Scathac's here complex came into play: she managed to convince us that the Goblins probably had at least one prisoner, from the whole wagon and dead horses thing, and that we, as good, law-abiding citizens, should go to their aid. I was inclined to argue against her, but she made a good point about Goblins often keeping treasure in their caves -- some of which may be magical.

We set out to follow them.

Turns out Emmeral is also pretty good at tracking down prey. He led us right to the Goblins' cave, though he seemed to have tried to off Scathac by dodging a trap and not telling her about it along our way. It was pretty funny to see her strung up from her foot, and even more so when Elwing tossed a dagger at the rope to get her down\dots\ from twenty feet above the path. Where she was hanging upside-down. Needless to say, she was not amused. She seemed mostly fine, though, so I was pretty amused.

The cave had a stream leading into it and a blind on the opposite side of the stream. We waded acorss the stream fine, but had to fight a few more Goblins hiding in the blind. They were, apparently, guarding the entrance, but they weren't quite aware enough to not be taken by surprise.

The cave, on the other hand, was less easy to manage.

We made our way in along the stream-side for only a few feet before a cavern opened to our right. Inside were three snarling wolves. I have no idea why Elwing insisted we should ignore them and continue moving on, but that's what we did. Further along, the path curved with a sort of landing across the stream on the other side. Just as we were approaching this, the stream quadrupled in size, a massive burst of water rushing for us. Scathac, I think, made the jump across to the landing, but the rest of us were swept away. It only took us barely out of the cave, but it certainly hurt -- both our bodies and our self-esteem.

The dead body Elwing had been carrying over his shoulder was confirmed dead this time, which at least means we didn't have to continue worrying about that. He didn't seem to torn up about it.

We tried to make our way back to the landing to meet with Scathac, but another burst of water scared us off. We jumped into the room with the wolves, and I was convinced we were done for, but Elwing managed to calm down the wolves enough so they didn't attack us. That was when I realized he could be useful after all, in the right circumstances. Scathac eventually re-joined us, since we realized there was a sort-of chimney chute in the back of the room that led deeper into the cave.

We climbed up the shoot (I need to get better at this, by the way, falling down due to lack of grip strength is \emph{not} a good feeling) and surveyed the new area: it seemed to be a mix of store-room and waiting-room. There were a few Goblins in there, a tamed wolf, and a Bugbear.

I'd never seen a Bugbear before then; they are terrifying. Massive and muscular, it looked like it could rip me in half without breaking a sweat. I decided to be clever, though, and used \textsc{Prestidigitation} to distract the Bugbear into leaving the room. A \textsc{Sending} convinced him that the Goblins were talking about him behind his back.

Well, sort of. It convinced him at first, which made him angry at the Goblins, which at least distracted them all enough to give us a bit of an advantage.

Our group took them by surprise and made short work of them. We managed to kill them all before any of them could run out an alert other Goblins, which I consider a win in our books. A well-timed \textsc{Fire Bolt} from me finished off the Bugbear. Unfortunately, nothing in the store-room held any value to us, nor was there any prisoners, so the monk had us continue onwards. She was insistent, that way.

We walked through the caves pretty slowly, after that, since we realized the benefit of surprising any Goblins that might want to attack us. We also started insisting Emmeral stay at the front of the group, for our own safety. That was pretty convenient for me, especially, since I didn't have to worry about getting attacked while I rained down hellfire on our enemies.

Or, well, something like that. Maybe ``hellfire'' is too strong of a word for it\dots\ but I digress.

The back of the cave was a large room, filled with a dozen Goblins. They were sitting around a campfire, screeching at each other in their awful language, with one of the Goblins repeatedly gesturing to his ripped off ear.

That was a bad sign, I know.

One of the Goblins sat on his own, on an out-cropping at the back of the room, overlooking the rest. A Dwarf was bound next to him, clearly unconcious. This, of course, was enough to make Scathac abandon all sense of surprise. She shouted at the head Goblin to let his prisoner go, which immediately alerted all of the Goblins that four adventures soaked in the blood of their brethren were tresspassing in thier cave.

Stupid heroes.

The Goblins around the fire stood and drew their weapons, the one on the outcropping held the Dwarf off of the ledge. He yelled at us in Common, asking what we wanted.

At this point, of course, the others were less sure about the whole yelling and alerting Goblins thing. I got elected our spokesman, which seems to continue to be a trend. Given how the conversation turned out, I'm a bit surprised the goody two-shoes of our group are alright with that but\dots\ well, I certainly don't mind.

I tried to convince the Goblin to give us the Dwarf, but we didn't really have anything to barter with. The Goblins seemed to be in fear of that Bugbear I killed earlier, though, so that gave me something to negotiate with: the head of the Bugbear in exchange for freeing the Dwarf. We were all set to make the trade when the conversatoin turned sour: the Goblin wanted more from us, a payment or something like that, and I was not willing in the slightest to actually lose anything to save this potentially dead, dying, or useless Dwarf. The Goblin said he'd drop the Dwarf if I didn't pay up and I dared him to follow through.

He did. The Dwarf probably wasn't too pleased by that, and Scathac and Elwing certainly weren't, but whatever. We ended up saving the Dwarf, so it all worked out fine.

The other Goblins launched at us, well, they attacked us -- some with swords and some with bows. The earless one especially came at us like a demon possessed. Emmeral put him down easily. Between the swords, fists, and magic of our little band of adventurers, the Goblins went down fast. Fast enough for Scathac to rush over and save the Dwarf before she died to her injuries. And the whole being dropped of a ledge thing.

Most interestingly, my magic did some things it had never done before: at one point, I launched a bunch of \textsc{Magic Missile}s at the Goblins hitting three of them in a single spell. I certainly wasn't expecting it, but it was something new and exciting that I knew I needed to figure out how to repeat. My magic has been doing that a lot, lately; I feel like I've been getting stronger over these past few days, learning how to do magic that years of exploring never taught me. The other have reported similar happenings, so it's not unique to me. Something about killing enemies and accomplishing our goals\dots\ it gives us all a lot of chances to practice our abilities.

Anyway, with the Goblins defeated, we trudged our way out of the cave, lugging the unconcious Dwarf with us. Elwing freed the wolves on our way out, which took absolutely none of us by surprise. We slept in the blind that night, since it was pretty much nightfall by the time we got out. I stayed up a bit thinking about my new magic, trying to remember the feeling of casting that spell so I might do it again, but eventually I fell into my trance as well.

That turned out to be a big mistake; we'd never even considered sleeping in shifts, which I'm kicking myself now for not suggesting.

\sleep
