% The Tenth Day of the Ninth Moon, Year 735

It's been two days now since we camped outside Vallaki, and so much has happened. I'd originally planned to write yesterday, but we were in jail, and I haven't had any downtime until now. Looks like that was the last I wrote, so I'll start there.

\section*{Vallaki}
We entered Vallaki yesterday morning, which seemed to have been a good choice given how much the city seems to hate anyone doing anything suspicious, like, say, entering the city in the dead of night. Kinnaeia looked exhausted, having not slept a wink but the rest of us were in decent spirits.

We walked through the city, taking in its dark and gloomy atmosphere, until we came across the town inn. Between our appearance and our race (elves are unique around here? seems odd, given Dusk Elves), we'd been getting enough odd looks that we decided to stay outside the inn, at first. Kinnaeia, Morianna, and Irena went in to inquire about rooms and ask about merchants in the city.

I have no idea why, but either Kinnaeia, Morianna, or both had the bright idea to pre-pay for their rooms. That turned out to not work so well for them! They also bought Irena a room, which she stayed in all day.

The rest of us made a bee-line to the town square, where Kinnaeia had been directed to by the innkeeper. This was the first time we say Izek, who I'm glad we are not on the bad side of. He has a strange arm that looks strong, and warped somehow. I would not want to get into a fight with him.

Izek was standing guard over the stockades, where some poor villager was locked. He was wearing a donkey mask; I wasn't sure at first whether that was the crime or the punishment, but it turns out this is how they punish people for ``not being happy enough''. The burgomaster is demented.

The weapons shop was convenient for Emmeral, at least. The rest of us didn't see anything useful for ourselves, but he bought a nice silver sword.

That seems like a great idea to me, by the way. I've read the lore, silver is supposed to be better against unholy creatures, which there seems to be a lot of in this country. All the better for us that he can deal with them.

The general store across the way, though, that was an absolutely wonderful find. Elwing and Emmeral bought flutes, but far more importantly: he had \emph{maps}! We each bought a map of Barovia, just to be sure, and I bought a map of Vallaki, as well. I had my head stuck in that map for ages after we found them, it feels wonderful to finally have some idea of where we are, where the other places we hear of are, and how to get between those places. I noticed the others speaking with Izek, but I didn't care much then and I don't really now; I was far too engrossed looking at the map.

I've been marking off my copy with the various tidbits of information we've heard in our travels -- it's quickly becoming well filled-in.

After a while, I noticed the conversation had stopped and Izek had left the area. Glancing down at my new map of Vallaki, I suggested we head east; there was another town swuare that way, and I was happy with our luck so far.

The area I'd directed us to wasn't at all the same as the first. It was more of a warehouse district, with several large, box-like buildings, the coffin-maker's shop, and Rictavio's caravan. Upon noticing that something within the caravan was growling, Elwing investigated a bit closer: as he tells it, the caravan contains a saber-toothed tiger, though both Rictavio and his guards simply described it as a normal lion. I'm inclined to trust the latter of those groups; we saw Elwing slip his hand through the bars, only to jump back a moment later as the beast jumped at him. Some druid he is.

The sounds of the beast clearly alerted Rictavio's guards, who ran out at us with their swords drawn and demanded Elwing back away from the animal. At this point, I should note, the rest of us were edging away from both the conflict and the clearly deranged druid.

The guards seemed content to let him go with a warning, though. In fact, they gave us some useful information that helped me get on Rictavio's good side, later; they mentioned that it was Rictavio that had captures the beast, and the Rictavio would be performing at the upcoming Festival of the Sun. A Festival I have no intention of intending, thank you very much.

At this point, Kinnaeia wanted to head back to the inn to drop off some of her equipment and check on Irena. The other's thought this was reasonable, though for some reason Elwing wanted us to look for a zoo (we vetoed that), so we headed back.

Kinnaeia, Morianna, and Elwing decided to head upstairs together, which left me and Emmeral to do some fact-finding. We entered the inn and both seemed to see the two sources of information at the same time: the barkeep, since barkeeps are always a good source of information, and the fanciful-looking folk in the corner. With a nod at each other, we split off; Emmeral to speak with the barkeep while I spoke with Rictavio.

Oh Rictavio, what an interesting man to find in a place such as Vallaki. He's an elf, though he didn't say it aloud and simply brought my attention to his ear (I attempted this again today, with a Dusk Elf, but this does not seem to be the way local Elves identify themselves, as I'd thought at this point). I got some interesting gossip out of him, once I impressed him enough with a simple \textsc{Prestidigitation} for him to ask if I would like to travel with him to perform. I'm not overly interested in being a performer, but this may give me a good excuse to travel around the area and explore. It's something I'll consider once he's left Vallaki.

Rictavio told me more about the upcoming festival, that it's dedicated to the Sun God of this area and is one of many festivals the burgomaster throws -- apparently he throws a festival that every citizen is required to attend every week in Vallaki. This festival is the day after tomorrow, exactly seven days after the previous one (the Festival of the Wolf Head, or something like that, which explains all the wolf heads on pikes outside). Rictavio will be performing at the festival, showing off the lion he'd recently captured and tamed. I mentioned Elwing's abilities with animals (still not sure he has these, other than the wolves from last week seeming to like him eventually) and informed him that I'd have the druid come speak with him later.

He also told me about the Vistani camp to the west. They seem to be the talk of the town; between all of us, we got directed that way nearly half a dozen times. In retrospect, this may have been the hands of fate guiding us; if we hadn't have gone to the Vistani camp, I wonder if we'd have heard the prophecy?

I said my goodbyes to Rictavio, promising to watch his performance if our ``wandering souls didn't guide my group of adventurers onwards'' (yes, I know), and caught the tail-end of Emmeral's chat with the barkeep.

Emmeral accomplished a bit more with his fact-finding than I did, not that I'm surprised; the barkeep gave him a good explanation of the surrounding area and some gossip which he passed along to me:
\begin{itemize}
\item the Vistani camp, again, and that it may be worth visiting,
\item Kresk, being the city to the west with the Abbey of Saint Markovia, was a safe haven,
\item Lake Baratok, which has an abandoned tower,
\item an abandoned city in a marsh to the south, which rebelled against Strahd, and
\item a haunted house to the south-west which used to house a dragon
\end{itemize}

When Elwing, Kinnaeia, and Morianna, joined us at the bar -- where we'd sat and ordered some wine while we discussed the area around us -- Emmeral passed along all the information he'd given to me (except the dragon, which makes sense to me -- I bet a dragon would guard some pretty interesting treasures, no point sharing those with people we don't trust). The fact that he kept some of the information between only us means he may have kept some information from me\dots\ but I'm not overly worried. The fact that he shared the gossip about the dragon tells me he thinks of me as a potential ally, which is enough for me.

I mentioned to Elwing that Rictavio was the one that had tamed the ``saber-toothed tiger''. Elwing got into a bit of a huff over that, and went over to give Rictavio a piece of his mind. Rictavio, though, just told some stories about his capture of both the lion and a pet monkey he once had, but had since given away to the toy-maker of Vallaki -- we actually saw that monkey performing outside the toy-maker's shop on our way out of the city today.

Rictavio's stories completely distracted Elwing; it was hilarious to watch him lose his train of thought like that. Maybe I should work on my story-telling ability so I can distract Elwing the next time he tries to do something stupid?

We headed west from there, continuing to explore the city, with the vague destination of Vallaki's church in mind. Outside, a young boy dug a grave -- Morianna chatted with him for a moment and we found there was no pressing need for the grave, rather someone would die eventually and would need a grave. Hearing the two talk made me depressed; both Morianna and the boy have no life to them, they feel utterly without motivation. I'd summarize their conversation as: ``the world sucks, doesn't it?'' ``yup''.

We entered the church, genuflecting in the same way other members of the conregation did, and listened to the sermon for a while. Speaking of depressing, the whole sermon was very somber: lots of talk about the state of everything being pretty sad. Not a happy place to be, that's for sure.

After the sermon, we talked with the priest for a moment. He offered sanctuary for Irena, which we gladly accepted on her behalf; and with that, our promise to Ismark was concluded. Upon noticing Kinnaeia, though, the priest tried to draft us into helping us with something. He pulled us to the side and spoke quietly about the protection around the church; a set of holy bones that apparently kept everyone safe.

If felt like needless superstition to me at first, though it turns out those bones were in fact magical, imbued with a protective spell. Either way, Kinnaeia was all for helping him and, since we didn't really have anything better to do, we agreed to go along with her. Emmeral was against it, but that seemed to have just motivated him to do the detective work faster.

The priest told us that he'd only mentioned the bones to one other soul; a young, scared altar boy, who we spoke with immediately. That altar boy directed us to the boy outside who we'd spoken to on our way in. The boy outside was less cooperative; it took Emmeral lifting him off the ground and interrogating him before the boy would talk.

It turns out, the boy had stolen the bones on behalf of the coffin-maker. Now annoyed that we had to walk clear across town, Emmeral took off in a huff and we followed after him.

\subsection*{Vampires in the Coffin-Maker's Coffins\dots\ I Get it Now}
Upon reaching the coffin-maker's shop, we knocked loudly. He told us to go away since he was closed, so I knocked again and demanded he let us in. Again, he told us to go away, so Emmeral broke down the door.

That was quite the sight for the coffin-maker, I bet. Two loud bashes on his door, then with the third the door simply shattered to pieces as Emmeral barreled through, the rest of us with raised weapons behind him. I'm sure I looked positively impressive, with the wisps of a \textsc{Fire Bolt} whirling around my finger tips. I must say, my ability to call upon at least a subset of my spells at will is becoming quite impressive. I swear I could hurl hundreds of \textsc{Fire Bolt}s in a day without care\dots\ it's intoxicating.

Anyway, the coffin maker was incredibly terrified (reasonable), so I played on his fear to demand he tell us where the bones were. Upon being informed the bones were upstairs, I ordered him to lead the way.

He lead us to the top of the stairs before falling into a quivering mess. At first, I was sure he was trying to trick us, but it turns out he, Labelas rest his soul, was simply terrified at the six vampires in the room we were about to enter. Apparently, it was at their command that he'd stolen the bones; they were planning on breaking into the church, we assume, and had used the coffin-maker as a scapegoat to remove the church's protections.

Sure enough, Emmeral was talented enough to sneak through the vampires' room without alerting them while Kinnaeia and I investigated the coffin-maker's bedroom next door. I felt confident there would be some back panel in his bedroom that would hide the bones -- Kinnaeia had cast \textsc{Detect Magic} and determined their approximate location as being behind the wall (I think I might have figured out how she did so? will need to think about it further) -- but I couldn't seem to find it. I did, though, find a colourful stone with a perfectly white circle on it -- something about it drew me to take it. I'm not sure why, but I've been meditating on it a while now. I'm almost convinced it's simply a bauble, but not quite yet.

Emmeral finished his investigation of the other room and let us know; the vampires were likely resting in the various boxes around the room, but did not seem to be awake. He told us he would need to move one of the coffins to get at the bones and that if a vampire was asleep in that coffin we'd have to fight.

Turns out he was pretty accurate in his report. He re-entered the room with the intention of taking any vampire within the coffin off-guard and yelled for help only seconds later. As we barelled into the room, we saw him wrapped up in a fight with the vampire that had previously been resting peacefully. Three other vampires were emerging from their coffins, and another two stirred in a couple nearby coffins.

We bounced into action: Kinnaeia cast \textsc{Turn Undead}, immediately causing the three vampires farthest from us to flee in terror, Elwing leaped onto the coffins of the slow vampires, turning into a bear to hold down the lids, and Morianna and I raced forward to help Emmeral.

This was the most terrifying fight I've been in in my life. Even the fight with Viktor, if you could even call it a fight, was nothing like the primal fear I felt upon seeing these vampires.

Somehow we actually managed to hold our own, at first. Between the awesome power of Kinnaeia turning aside the undead, Emmeral's flashing swords, and my growing magic, we even managed to kill two of them. Whether they stay dead or not, mind, I have no idea; we need to figure out how to actually fight vampires, given this seems to be a pretty common thing around here.

After we killed the first one, I ran to grab the bones. That they were hidden in a chest with a false bottom almost confused me, but I figured it out.

After I returned to the fight with the bones, the vampires started going after me a bit more. My mage armor was enough to protect me, but only barely. Once Kinnaeia's \textsc{Turn Undead} wore off, the fight quickly turned against us.

We fought for barely a minute before we realized it was hopeless. Emmeral and I seemed to have the same idea at the same time; just as I started making plans to escape rom the fight, he asked me to throw him the bones to ``distract the vampires by running away''. No way that was gonna happen.

Seconds after I yelled a ``no'' to him, Elwing let out a \textsc{Thunder Wave}. It did a lot\dots\ to us. The vampires seemed unaffected, but Morianna and Emmeral were visibly winded. Unperturbed by this, Elwing cast another \textsc{Thunder Wave}, this time knocking Morianna and Emmeral completely unconscious. It knocked the vampires away from a moment, but it still helped them far more than it helped us.

With half the group nearly killed by Elwing, I began to flee the room. Let the others get killed, eat, or vampirized, I didn't care, I just wanted to survive. And to get away from Elwing, whose motivations I can not figure out whatsoever. Is he trying to kill us? I have no idea. He seems to be on our side for the most part, even seems to genuinely want to be helpful, but then he goes and does things like that. Maybe he is just terrible at thinking on his feet; again, maybe I need to ingratiate myself with him enough to have him obey me in combat.

Either way, as I fled the room, Elwing did an about-face again. Refusing to leave without the others, he transformed again into a bear (looks like he can't use magic while transformed?), lifted Morianna in his mouth, had Kinnaeia jump on his back, and carry Emmeral. It was\dots\ impressive\dots\ how he managed to save the other three simply by being a bear.

I need to figure out how to do that.

As we raced out of the room, groups of guards raced in to hold back the vampires. We never did find out whether they won or lost, whether the vampires fled or killed them, but at least one of the guards survived to inform the burgomaster or our ``complicity in bring vampires into Vallaki''. Fool.

With the guards holding the vampires away from us, we managed to escape. Still, if it hadn't been for the perfect circumstances there, I'd be dead right now. Not a good thought at all.

We raced back to the church. For some idiotic reason, Kinnaeia began to yell at the priest that we'd recovered the bones well within the earshot of the congregation that the priest had explicitly told us the bones were being kept a secret from. She seemed abashed enough when the priest called her on it, though still tried to insist he pry up the floorboards in front of the congregation to replace the bones.

It's as if I'm surrounded by idiots sometimes, I don't understand it.

The priest thanked us, though he appeared as annoyed as I was at Kinnaeia's outburst. He insisted we ask for a reward from the burgomaster's wife (that's never going to happen now) and we told him we'd bring Irena to the church first, to see her safely away, then head on to see the burgomaster's wife.

\subsection*{Arrested}
We returned to the inn and into Irena's room, completely oblivious to the guards organizing around us. We should have kept a lookout, but no; we all walked into Irena's room (without knocking -- Kinnaeia got stabbed by a terrified Irena) and, when we exited, a dozen guards were set to arrest us for ``causing a disturbance''.

They refused to respond to questions as to why we were being arrested, other than informing us the burgomaster requested it. That's about the point when I realized the reward for protecting the bones would not be happening.

Emmeral resisted arrest; when the guards demanded we disarm, he simply crouched with his swords on the ground. Following his idea, I dropped my as-of-yet unused crossbow and declared myself ``unarmed''. A second later, Emmeral sprinted away from us, weapons still in hand, attempting to escape. The guards were ready for this, though, and tripped him, kicked away his weapons, and manacled us all together. Maybe they'd have done that anyway, but it certainly wasn't pleasant. I'm not opposed to Emmeral having tried to escape, but I wonder if he'd have come back to free us if he'd succeeded? I feel confident my magic would have saved me, but it would be good to know how much he is on my side.

The guards brought us to a jail underneath the burgomaster's house with an incredibly convenient six unused cells next to each other. Stripsearching us for all our equipment, they threw us in the cells like vermin, no matter how I tried to convince them to let us speak with the burgomaster immediately. Izek stood watch over us all night, not even moving a muscle, after informing us the burgomaster would speak with us at his convenience the next day. Nightmarish. They took everything from us; my focus, my spellbook, this journal, everyone else's equipment, everything.

Elwing again confused me that night by being incredibly useful; after discussing with each other (in Elvish, obviously. We didn't want Izek to overhear and understand -- Irena, Morianna, and Kinnaeia slept through this since they don't speak Elvish. Maybe we should teach them?) about what we should do (escape together? have one of us escape? stay?), Elwing went off as a spider to gather intelligence.

He went off twice; once to speak with the only other occupant of the prison and once to explore the burgomaster's house. Both times, we arranged it such that he went off between the changing of the guard so he would not be noticed.

He didn't find much out from exploring, other than apparently seeing a saddened maid cleaning up the dining room. Knowing she was sad was useful later, but it certainly didn't help us out that night.

From the other prisoner, we learned a lot, though.

As relayed by Elwing, the other prisoner was happy to stay in prison rather than attempt to escape; his punishment would simply be a day of embarassment in the stocks while wearing a donkey's head. He had no idea how we would be punished, but offered some valuable insight into the burgomaster.

The burgomaster is a proud, arrogant man, insistent that Vallaki is a safe haven where ``all is well'', all the time. Citizens who aren't ``happy enough'' are arrested by his men, and no negative talk is permitted. This gave me enough knowledge to convince the burgomaster to free us the next day\dots\ I'm sure I could have managed without Elwing's intelligence, but he was\dots\ helpful, I suppose.

Then again, he did have the stupid idea of stealing servant clothing and disguising himself. How the burgomaster would not realize that he now had five servants, one of which he didn't recognize\dots\ Elwing certainly doesn't have the best of ideas.

One of the most annoying parts of getting arrested was how much time it wasted. I could have spent several hours that night examining the spellbook, updating this journal\dots\ as is, I'm sure I've forgotten something that happened that day, what with all that's happened, and I can't shake the feeling that it might be important. At least tonight I can study the spellbook more, but I feel like I'm so close to deciphering it. If I could have spent hours on it last night, maybe yesterday would have turned out a bit better.

\sleep
