I've never been one to write too much; that's something wizards do, and I'm certainly no wizard. Wizards study for years to learn how to perform even the simplest of magic, but for me\dots\ magic just comes naturally. I don't mean to say it's easy for me -- it's certainly been far more disastrous than any wizard I've ever heard of -- but it's not a studied skill. So when I say I'm not much of a writer, I mean it.

But in my frankly absurd past few days I've come across a book. A genuine Wizard's Spellbook. And I'm excited, because I know it contains some of the secrets of magic that I've been searching for.

It's been tough going, though; even with the advantage of, you know, actually being able to cast spells already and so not learning everything from scratch, I've been staring at this book for what feels like forever (three days? has it only been that long?) with no progress. Just last night, I felt like I'd finally made a breakthrough\dots\ but still, nothing in this book is working for me.

Maybe I should start at the beginning? I've heard people with journals like to do that. And I do have a good place to start, what with this whole Barovia thing.

This seems like a good idea: let's start here.

\section*{Dear Diary}
My name is Kal'Thruum, and I am a sorcerer. Not a cleric, like Kinnaeia, or a warlock, or a wizard. My magic comes naturally, wildly, and often unexpectedly. It's always bothered me more than anything that my magic, though I can at least somewhat control it by now, doesn't quite work as I want it to -- spells sometimes fizzle out or do the complete wrong thing for me. To that end, I've spent the best part of my life searching for answers: how does magic work? Where does it come from? Why am I the way I am?

I left home early, for a half-elf. A bit early for a human, even. As soon as I discovered my magic, I couldn't hold myself back -- I needed to know.

I went off exploring. I can't say I had a destination in mind, or even a plan, but I was confident in my abilities. Not the most intelligent move I've ever made, but then I never claimed to be the smartest or wisest person out there.

It took about a year before my first experience with magic. I'd been living on the streets, living off the charity of strangers. I'd always been pretty good at convincing people to help me out when I needed to, and that saw me through that first year. I was living in Neverwinter at the time, having wandered west from Llorkh. Neverwinter was a lovely town for what I was looking for: the Blacklake district was filled with the rich and guillible and Beggar's Nest had a graveyard which had some propensity for the undead. Zomebies, but no one quite knew why they appeared. I had no interest in fighting the zombies, mind, especially back then when I could barely cast a consistent \textsc{Fire Bolt}, but the source of the magic animating them in the first place had me intrigued.

Eventually, I did in fact find this source: an artifact in the graveyard, hidden in one of the crypts, which positively pulsed with necromancy. It was a huge tablet, covering an entire wall of the crypt, and inscribed with the most curious of letters: Iokharic. Before this, I'd had no experience with any languages other than Elvish and, of course, Common, but the wall filled me with so much excitement. Here it was! Real magic, secrets that no one had seen in years, written in a foreign language as old as time itself.

Once I'd identified the language, I searched for a way of learning it. Eventually, I came across an elderly elf by the name of Meldanen -- he took great pleasure in seeing the tricks that I could accomplish with a simple \textsc{Prestidigitation} and clearly had more money than sense. But, he could speak the language of dragons, and gladly passed on the knowledge in exchange for a simple fireworks show here and there. The poor fool didn't know nor care that the tricks cost me nothing, and were the simplest trick of all -- something which greatly benefitted me, since they were about the only spell I could safely cast without danger.

After far too much study, I could finally understand and speak the language of dragons. I returned to the crypt, eager to understand its mysteries\dots\ and was dissappointed to find nothing of interest. The writing seemed to be nothing more than an explanation of why the artifact was there -- apparently, in years past, an archeologist was exploring the crypts having been convinced there was someone of importance who'd been buried there, forgotten. Worried about being interrupted by the townsfolk, he designed this artifact to scare them away. Conveniently, there was an explanation as to how to disable the effect: I'd always planned to return to the city at some point to convince some rich fool to pay me to turn it off. With our sudden departure of the Forgotten Realms, this may never happen\dots\ but I'm hopeful.

The method, by the way, is to chant in front of the tablet: \textsl{Si relgr acht wer vers di wer darastrix ekess sulta wer chikohk di tisultair timarfedelom}.

Disheartened, I searched for anything else in the crypt that could sate my hunger for knowledge. I am glad my search got so frantic, since in my frenzy I broke through the coffin next to the tablet. Inside was, not a dead body as I would have expected, but a diary in a language I could not recognize. The only part of the diary I've been able to identify is a marking on the back page: it is the same as a marking on the tablet that I consider the author's signature. My conclusion, then, is that this diary is some journal written by the original archeologist.

I returned again to the city proper to find anyone who would recognize the language, but was met with nothing but blank stares. Eventually, I will decipher the diary\dots\ but I suppose I have more important things on my plate for now.

Frusterated with my lack of progress, I decided to search for magic elsewhere. My destination this time was the Neverwinter Woods, which I had been informed held some mysteries: strangers visiting Neverwinter had told stories for years of shapeshifters (obviously druids), tricksters (clearly Fae), and ghosts (not sure about this one). I'd of course been warned away from Fae by father, but then, my mother never seemed so afraid, and I always took after her more than him. They, I knew, held some of the secrets of magic, being so closely tied to it themselves.

I bartered an Adventurer's Pack out of a merchant in the City Core in exchange for a few days of help around the shop and set out. I even convinced him to donate a few simple daggers to my cause, since I was ``so frail and unable to protect myself from the ghosts''.

The forest was different than I'd expected. I never did manage to find the Fae, but I did find something better: the trees themselves were magical, able to think and speak just like you or I. They have their own language, even: Sylvanic, which I eventually picked up to understand what the trees knew about the magic of the forest. To summarize: not much. Trees can speak, sure, but they don't think the same way as we do, they can be hardly made to care for the whims of a half-elf. In fact, they don't really provide much value at all: why do they need a language if they never really say anything?

Goddamn trees.

They did direct me to a Fae camp, though it wasn't quite intentional on their part. The Fae don't interfere with the trees like men do, but they make enough of an impact that it\dots\ let's say it comes up in conversation among trees.

Fae are weird creatures. Flighty, vain, nearly as hard to speak with as trees. They speak the same language, for the most part, depending on which Fae they are. Sylvan, Common\dots\ some of them even speak Elvish. They seemed to take delight in tricking me, casting their illusions to fool me into believing the patently absurd. My willingness to go along with their jokes and my ability to speak their languages somewhat enamoured them of me, though, and I certainly learned some of what I set out to accomplish. They gifted upon me an arcane focus made of a glowing crystal, one of my proudest possessions. They informed me of its dual uses: doubling for the physical components of any more complicated spell and increasing my virilence. I'm not convinced of the latter, but the former has certainly been proven time and time again.

My ability to cast \textsc{Dancing Lights} was never something I understood: it seemed to be tied to nothing in particular and would sometimes work and sometimes have no effect with no readily apparent reason. It turns out, though, the reason was the proximity of wychwood; somehow, I can summon all the arcane energy necessary to create these lights -- but only by consuming the energy of nearby wychwood. My previous ``random'' abilities, then, were based on my proximity to wychwood, something I'd never even considered could affect the magic.

With the focus, though, I can cast \textsc{Dancing Lights} with certainty. The feeling upon realizing I had gained some modicum of control over my powers\dots\ it was a good feeling.

I stayed with the Fae for a while. It felt like a couple years, but I know the Fae have a way of tricking time itself into disabeying the whims of the real world. For all I know, hundreds of years have passed since I first entered that forest. Events have contrived themselves to prevent me from returning to anywhere I'd been before\dots\ well, you'll see. I can't say I care too much, after all, I don't seem to be suffering any ill effects.

Wow, I've certainly rambled for longer than I expected. I wrote earlier that I had a beginning in mind\dots\ well, I haven't gotten to that part yet. That happened a few days after leaving the Fae, about five days ago now.
