\section*{Prophecy}
Four of glyphs, trader, is ``history, knowledge of the ancient, to better understand our enemy'', says ``look to the wizard of wines, in wood and sand treasure hides''. Five of stars, enchanter, is ``powerful force for good and protection, a holy symbol of great hope'', says ``kneeling woman, a rose of great beauty plucked too soon, the master of the marsh knows of whom I speak''. One of stars, avenger, is ``power and strength, a weapon of vengeance, a sword of sunlight'', says ``the treasure lies in a dragon's house, in hands once clean and now corrupted''. The beast card says ``a werewolf holds a secret hatred for your enemy, use her hatred to your advantage'' is ``this card sheds light on one who will help you greatly in the battle against darkness''. The broken one says ``he haunts the tomb of the man he envied above all'' and is ``there is a creature of darkness who's powers are beyond mortality, this card will lead you to him''.

``Each of you have a great and terrible destiny, a great and terrible challenge''.

Ok, what was that? Arabelle just walked in with magic billowing out of her eyes to deliver a prophecy and I have no idea what is going on. First things first, we need to figure out who this enemy is. I'd assume Strahd, but we're not enemies with him, are we? I know I have no current intention of fighting against him, but maybe the others do? Would Arabelle, or whatever other-wordly force took over her body just now, know that?

Let's assume it's Strahd, that's the obvious answer. I guess it might be someone or something we don't know about yet, but\dots\ no, my gut tells me it's Strahd.

Alright, then let's figure out what this would mean. I'm pretty sure the trader refers to the Wizard of Wines winery, like is on the Vistani barrels of wine, and there's some sort of treasure in ``wood and sand'' near the winery that will tell us more about his past. The enchanter seems to talk about some sort of protective holy symbol; there's some young woman protecting it, maybe? I know that abandoned town to the south was near a marsh, the ``master'' of that place -- whatever that means -- probably could point us to this woman?

The avenger card is pretty straight-forward: a sword that can kill vampires, in the haunted house where the dragon was seen, and that there's someone there who is corrupted and currently holds the sword. The beast card seems to point to an ally, a werewolf woman that would fight against Strahd. And the broken one is about some immortal, powerful being in a tomb of some man he envied? That last one seems the most confusing to me, but I'll keep my eyes out for tombs. Or, uh, people with ``powers beyond mortality'', but I would be watching out for that anyway.

And as for the last line\dots\ I don't know. I never thought I had some ``great and terrible destiny'', but\dots\ well, I don't know what to think.

\pause

Ok, now that I've written that down, I just need to not think about it for a bit. Maybe it'll make more sense after a while, I don't know. The others are trancing or sleeping or passed out, so clearly they're thinking the same thing. I don't know.
